### 2014-11-09 Abstract.tex
stuff

We present a new framework for modular verification of hardware designs
in the style of the Bluespec language.  That is, we formalize the idea of
components in a hardware design, with well-defined input and output channels;
and we show how to specify and verify components individually, with
machine-checked proofs in the Coq proof assistant.  In doing so, we import
to the world of hardware those modular reasoning principles that are
well-known in software verification.  As a demonstration, we verify a
fairly realistic implementation of a multicore shared-memory system.  The
main modularity split is between a memory-system component and a processor
component.  Both include nontrivial optimizations, with the memory system
employing an arbitrary hierarchy of cache nodes that communicate with each
other concurrently, and with the processor doing speculative execution of
many concurrent read operations.  Nonetheless, we prove that the combined
system implements sequential consistency.  One interesting aspect of the proof
is that the memory component is well-specified to be reused for systems
implementing a variety of weak memory models.  To our knowledge, our proof
of that memory system is the first machine verification of a cache-coherence
protocol parameterized over an arbitrary cache hierarchy, and our full-system
proof is the first machine verification of sequential consistency for a
multicore hardware design that includes memory and processors.

>>>>

We present a new framework for modular verification of hardware designs in the
style of the Bluespec language.  That is, we formalize the idea of components
in a hardware design, with well-defined input and output channels; and we show
how to specify and verify components individually, with machine-checked proofs
in the Coq proof assistant.  As a demonstration, we verify a fairly realistic
implementation of a multicore shared-memory system consisting of two types of
components:  memory-system and processor.  Both components include nontrivial
optimizations, with the memory system employing an arbitrary hierarchy of cache
nodes that communicate with each other concurrently, and with the processor
doing speculative execution of many concurrent read operations.  Nonetheless,
we prove that the combined system implements sequential consistency.  One
interesting aspect of the proof is that the memory component is well-specified
to be reusable for systems implementing a variety of weak memory models.  To
our knowledge, our memory-system system proof is the first machine verification
of a cache-coherence protocol parameterized over an arbitrary cache hierarchy,
and our full-system proof is the first machine verification of sequential
consistency for a multicore hardware design that includes caches and
speculative processors.

### 2014-11-09 Introduction.tex
stuff

A modern high-performance, cache-coherent, distributed-memory hardware system is
inherently complex. Such a system implements memory models like sequential
consistency (SC)~\cite{lamport1979make}, Total Store Order (TSO), and
Relaxed Memory Order (RMO)~\cite{weaver1994sparc}, each
of which characterizes how loads and stores may be issued by a processor and
how the shared-memory component responds to these requests.  These systems by
their nature are highly concurrent and nondeterministic.

The goal of this work is to provide a framework for full verification of such
complex hardware systems. Hardware designers already recognize the challenge of
implementing these systems, to such an extent that they have already become some
of the most serious real-world adopters of formal methods.  Hardware
verification is dominated by model-checking; for instance, processor
verification ~\cite{burch1994automatic, mcmillan1998verification} and 
more recently, Intel's execution cluster verification~\cite{kaivola2009replacing}.
Many abstraction techniques are
used to reduce designs to finite state spaces, which can be explored
exhaustively.
 There are limits to the construction of sound abstractions, so
verifications of, e.g., cache-coherence protocols have almost always treated
systems with concrete topologies, involving particular finite numbers of caches
and processors. For instance, explicit-state model checking tools like
Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
addresses, and one bit of data. Symbolic model checking techniques have fared
slightly better: McMillan \etal{} have verified a 2-level MSI protocol based on
the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp, Mccomp, mcc}, \etc{}) scale
the approach further, but they are still unable to handle realistic
hierarchical protocols. We might hope to achieve higher assurance and
understanding of design ideas by verifying \emph{infinite families} of hardware
designs, which resist reduction to finite-state models.  A potential related
benefit is reducing the computational resources required to formally verify a
new design in the family, applying a more general theorem without much new
analysis.

@@

%\xxx{\textbf{Murali's version in red} This works introduces an interface enabling true modular implementation and
%verification of shared memory systems. We want our modular system to permit
%\emph{black-box refinement}, that is, nothing beyond the interface contract can
%be assumed about other modules.  That is, we should be able to describe a
%modular decomposition where we may replace either the processor nodes, the
%memory with locally verified alternatives without changing the correctness of
%our whole system.  To this end, we identify a key property called \emph{store
%atomicity} shared by the memory component of every memory model. This property
%effectively restricts the behavior of the memory sub-system to that of an
%\emph{atomic instantaneous memory}. We prove that this property is in fact what
%is guaranteed using the cache coherence protocols in a multi-processor system
%containing a hierarchy of caches, enabling replacement of the cache hierarchy
%with an atomic instantaneous memory. Similarly, on the processor side, we
%define what is meant to be a correct single processor. These invariants about
%the processor and the memory sub-system remains constant across memory models.
%We show a concrete instance of this framework by proving that a highly
%speculative multi-processor system connected to a coherent cache hierarchy
%implements the sequential consistency memory model.}

In this paper, we introduce \textbf{a new approach to modular,
  deductive verification of hardware designs} at the highest level of
rigor, backed by proofs using the Coq proof assistant.  As a
challenging case study, we focus on \textbf{verifying that a
  particular infinite family of multicore, shared-memory systems
  implements sequential consistency}.  The proof is parametrized over
an unknown number of processors connected to an arbitrary memory
hierarchy with an unknown number of caches in an unknown number of
layers (e.g., L1, L2).  In our design, processors and memory systems
both independently employ intricate optimizations that exploit
opportunities for parallelism.  We are able to prove that each of
those two main components still implements strong enough semantics to
support SC, and then we compose those theorems into a result for the
full system.  Either component may be optimized further without
requiring any changes to the implementation, specification, or proof
of the other.

To structure our hardware verification framework, we import
\emph{labeled transition systems (LTSes)}, a well-studied approach from the
world of concurrent software modeling.  As our unified notion of
specification and proof, we adopt \emph{trace refinement}, which
captures when one concurrent system can only produce observable IO
behaviors that another system could also produce.  Each of our
realistic hardware components is associated with a simpler ``reference
implementation'' that serves as a specification, and we prove that
each realistic component refines its spec.  Thereafter, it is sound to
use the simpler spec component in reasoning about the behavior of the
system.

Each key component of our case-study system has a simple spec: the
ideal memory responds instantly and atomically to each load or store
request, and the ideal processor executes instructions in order with
no speculation.  The optimized implementation of each is much more
complicated: the cache-hierarchy memory system potentially allows
every cache to be busy simultaneously computing and sending messages,
and the optimized speculative processor sends multiple load requests
to the memory at once, without waiting for responses first.  The meat
of the verification is in showing that each optimized component
refines its simpler counterpart.  After each such proof, we may in
effect substitute the optimized version for the simple version in a
black-box way within a design, with a full guarantee of soundness.

LTSes as hardware descriptions are an established
idea~\cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are
compilers that convert LTSes into efficient hardware.  Our work is
based on the Bluespec language~\cite{BSV:LangRef, Bluespec:TFRG},
which served as the model for the formalism of this paper.  Bluespec
models hardware components as atomic rules of a transition system over
state elements, and there is a commercial compiler from such code into
register transfer-level circuits (i.e., Verilog code) with competitive
performance.  Our cache-coherent memory system is directly
transliterated from a Bluespec
implementation~\cite{DNA:CoherenceImplementation} used to
implement a multi-processor PowerPC system~\cite{Khan:PowerPc}. The
hardware synthesized from that implementation is rather efficient:
an 8-core system can run 55 million instructions per second on
the BEE FPGA board\cite{}, with a 2-level cache hierarchy. Figure \ref{both} gives an example
of an LTS inference rule transliterated directly from an actual Bluespec rule used
in the PowerPC system.

@@

\textbf{endrule}
\end{boxedminipage}
\begin{boxedminipage}{\columnwidth}
\small
\inference[]
{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
{\ioxe{$M_c$}
{\dt, \ch, \s,\\ \dst, \wt, \dwt,\\ \inp, \outp}
{\dt, \ch[ (c, p, \req) \coloneqq(a, \s(c, a),x)\\
\;\;\uplus \ch(c,p,\req)], \s, \dst,\\
\wt[(c,a)\coloneqq x], \dwt,\inp, \outp}{}}
\end{boxedminipage}
\caption{Rule in Bluespec and its corresponding LTS transliteration}
\label{both}
\end{figure}

@@

\item A general \emph{modular verification methodology} applicable to hardware
  algorithms, based on \emph{labeled transition systems}.
We are able to verify
  the several components of a modern processor 
%\adam{(Figure~\ref{zoom} OK to have a forward reference to this figure?}
against general formal interfaces,
  enabling mixing and matching of different realizations of each interface,
  without doing any new proofs that peek beneath abstraction boundaries.

@@

\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce our
flavor of the labeled transition systems formalism, including a definition of
trace refinement. In Section~\ref{sec:store-atomicity}, we show a generic
decomposition of any multi-processor system independently of the memory model
that it implements, and we discuss the store-atomicity property of the memory
sub-component. In Section~\ref{sec:sc} we
give a simple formal model of sequential consistency.  The
following sections refine the two main subcomponents from Figure~\ref{zoom}.
Section~\ref{sec:ooo} discusses definition and verification of a speculative
processor model, and Section~\ref{sec:cc}
defines and proves our hierarchical cache-coherence protocol.  Finally, in
Section~\ref{sec:finalresult} we show the whole-system modular proof of our
complex system, finishing with discussion of related work and other conclusions in
Sections~\ref{relatedWork} and \ref{sec:conclusion}.

>>>>

A modern high-performance, cache-coherent, distributed-memory hardware system
is inherently complex. Such a system implements memory models like sequential
consistency (SC)~\cite{lamport1979make}, Total Store Order (TSO), and Relaxed
Memory Order (RMO)~\cite{weaver1994sparc}, each of which characterizes how
loads and stores may be issued by a processor and how the shared-memory
component responds to these requests.  These systems by their nature are highly
concurrent and nondeterministic.  The goal of this work is to provide a
framework for full verification of such complex hardware systems.

@@

In this paper, we introduce \emph{a new approach to modular, deductive
verification of hardware designs} at the highest level of rigor, backed by
proofs using the Coq proof assistant.  As a challenging case study, we focus on
\textbf{verifying that a particular infinite family of multicore, shared-memory
systems implements sequential consistency}.  The proof is parametrized over an
unknown number of processors connected to an arbitrary memory hierarchy with an
unknown number of caches in an unknown number of layers (e.g., L1, L2).  In our
design, processors and memory systems both independently employ intricate
optimizations that exploit opportunities for parallelism.  We are able to prove
that each of those two main components still provides strong enough guarantees
to support SC, and then we compose those theorems into a result for the full
system.  Either component may be optimized further without requiring any
changes to the implementation, specification, or proof of the other.

To structure our hardware verification framework, we use \emph{labeled
transition systems (LTSes)}, a well-studied approach in the world of
concurrent software modeling.  As our unified notion of specification and
proof, we adopt \emph{trace refinement}, which captures when one concurrent
system can only produce observable IO behaviors that another system could also
produce.  Each of our realistic hardware components is associated with a
simpler ``reference implementation'' that serves as a specification, and we
prove that each realistic component refines its spec.  Thereafter, it is sound
to use the simpler spec component in reasoning about the behavior of the
system.

Each key component of our case-study system has a simple spec: the ideal memory
responds instantly and atomically to each load or store request, and the ideal
processor executes instructions in order with no speculation.  The optimized
implementation of each is much more complicated: the cache-hierarchy memory
system potentially allows every cache to be busy simultaneously computing and
sending messages, and the optimized speculative processor sends multiple load
requests to the memory concurrently, without waiting for responses first.  The meat
of the verification is in showing that each optimized component refines its
simpler counterpart.  After each such proof, we may in effect substitute the
optimized version for the simple version in a black-box way within a design,
with a full guarantee of soundness.

LTSes as hardware descriptions are an established
idea~\cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are compilers that
convert LTSes into efficient hardware.  Our work is based on the Bluespec
language~\cite{BSV:LangRef, Bluespec:TFRG}, which served as the model for the
formalism of this paper.  Bluespec models hardware components as atomic rules
of a transition system over state elements, and there is a commercial compiler
to synthesize such codes into register transfer-level circuits (i.e., Verilog code) with
competitive performance.  Our cache-coherent memory system is directly
transliterated from a Bluespec
implementation~\cite{DNA:CoherenceImplementation} used to implement a
cycle-accurate simulator for a multi-processor PowerPC system~\cite{Khan:PowerPc}. The hardware synthesized
from that implementation is rather efficient: an 8-core system with a 2-level
cache hierarchy can run 55 million instructions per second on the BEE FPGA
board\cite{bee2}. Figure \ref{both} gives an example of an LTS inference rule
transliterated directly from an actual Bluespec rule used in the PowerPC
system. In Bluespec notation, only those state elements whose value changes are
mentioned on the RHS of a rule; in
contrast LTSes employ a more functional notation. (The meaning of this rule
will become clear in Section \ref{sec:caches})

@@

\textbf{endrule}
\end{boxedminipage}
\begin{boxedminipage}{\columnwidth}
\small
\inference[]
{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
{\ioxe{$M_c$}
{\dt, \ch, \s,\\ \dst, \wt, \dwt,\\ \inp, \outp}
{\dt, \ch[ (c, p, \req) \coloneqq(a, \s(c, a),x)\\
\;\;\uplus \ch(c,p,\req)], \s, \dst,\\
\wt[(c,a)\coloneqq x], \dwt,\inp, \outp}{}}
\end{boxedminipage}
\caption{Bluespec rule and its corresponding LTS rule}
\label{both}
\end{figure}

@@

\item A general \emph{modular verification methodology} applicable to hardware
algorithms, based on \emph{labeled transition systems}.  We are able to verify
the several components of a modern processor against general formal interfaces,
enabling mixing and matching of different realizations of each interface,
without doing any new proofs that peek beneath abstraction boundaries.

@@

\paragraph{Paper Organization:} We begin with a discussion of related work in
Section~\ref{relatedWork}. In Section~\ref{sec:lts} we introduce our flavor of
the labeled transition systems formalism, including a definition of trace
refinement. In Section~\ref{sec:store-atomicity}, we show a generic
decomposition of any multi-processor system independently of the memory model
that it implements, and we discuss the store-atomicity property of the memory
sub-component. In Section~\ref{sec:sc} we give a simple formal model of
sequential consistency.  The following sections refine the two main
subcomponents of our multi-processor system.  Section~\ref{sec:ooo} discusses
definition and verification of a speculative processor model, and
Section~\ref{sec:cc} defines and proves our hierarchical cache-coherence
protocol.  Finally, in Section~\ref{sec:finalresult} we show the whole-system
modular proof of our complex system, and end with some conclusions in
Sections~\ref{sec:conclusion}.

### 2014-11-09 RelatedWork.tex
stuff

%Formal verification has been used to verify multiprocessor
%systems, especially with model checking
%\cite{burch1994automatic,mcmillan1998verification}.

%More recently, Intel's execution cluster (\ie{} the ALU and its
%interface) has been formally verified using model checking
%\cite{kaivola2009replacing}.

>>>>

Hardware verification is dominated by model-checking; for instance, processor
verification ~\cite{burch1994automatic, mcmillan1998verification} and more
recently, Intel's execution cluster verification~\cite{kaivola2009replacing}.
Many abstraction techniques are used to reduce designs to finite state spaces,
which can be explored exhaustively. There are limits to the construction of
sound abstractions, so verifications of, e.g., cache-coherence protocols have
almost always treated systems with concrete topologies, involving particular
finite numbers of caches and processors. For instance, explicit-state model
checking tools like Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to
handle small and simplified systems, \eg{} a single-level cache hierarchy with
three CPUs, two addresses, and one bit of data. Symbolic model checking
techniques have fared better: McMillan \etal{} have verified a 2-level MSI
protocol based on the Gigamax distributed multiprocessor using
SMV~\cite{gigamax}. Optimizations on these techniques (\eg{} partial order
reduction~\cite{part}, symmetry reduction~\cite{sym1, sym2}, compositional
reasoning~\cite{comp, Mccomp, mcc}, \etc{}) scale the approach further, but
they are still unable to handle realistic hierarchical protocols. We might hope
to achieve higher assurance and understanding of design ideas by verifying
\emph{infinite families} of hardware designs, which resist reduction to
finite-state models.  A potential related benefit is reducing the computational
resources required to formally verify a new design in the family, applying a
more general theorem without much new analysis.
